---
title: "Lab 7  -- Polynomial Models"
author: "Philipp Broniecki and Lucas Leemann -- Machine Learning 1K"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(global.par = TRUE)
```

##### (based on James et al. 2013)

We begin by loading that `ISLR` package and attaching to the Wage dataset that we will be using throughout this exercise.

```{r, eval=FALSE}
# clear workspace, load ISLR, attach wage data set
rm(list=ls())
library(ISLR)
attach(Wage)
?Wage # codebook
```

```{r, include=FALSE}
# clear workspace, load ISLR, attach wage data set
rm(list=ls())
library(ISLR)
attach(Wage)
```

### Polynomial Regression

Let's fit a linear model to predict `wage` with a forth-degree polynomial using the `poly()` function.

```{r}
# linear regression on wage, with age up to a 4th degree polynomial
fit <- lm(wage ~ poly(age, 4), data = Wage)
coef(summary(fit))
```

We can also obtain raw instead of orthogonal polynomials by passing the `raw = TRUE` argument to `poly()`. The coefficients will change the fit should be largely unaffected.

```{r}
fit2 <- lm(wage ~ poly(age, 4, raw = TRUE), data = Wage)
coef(summary(fit2))
```

There are several ways to specify polynomials. These are, however a little less convenient.

```{r}
fit2a <- lm(wage ~ age + I(age^2) + I(age^3) + I(age^4), data = Wage)
coef(fit2a)
```

A more compact version of the same example uses `cbind()` and eliminates the need to wrap each term in `I()`. The output is less readable though.

```{r}
fit2b <- lm(wage ~ cbind(age, age^2, age^3, age^4), data = Wage)
coef(fit2b)
```

We can create an age grid for the targeted values of the prediction and pass the grid to `predict()`. We can set the argument `se=TRUE` in the `predict()` function which will return a list that includes standard errors of the outcome. We can use these to an upper and lower bound of our estimate of $y$.

```{r}
# minimum and maximum values of age variable
agelims <- range(age) 
age.grid <- seq(from = agelims[1], to = agelims[2])
# se=TRUE returns standard errors
preds <- predict(fit, newdata = list(age = age.grid), se = TRUE)
# confidence intervals as estimate + and - 2 standard deviations
se.bands <- cbind(preds$fit + 2 * preds$se.fit, preds$fit - 2 * preds$se.fit)
```

We can plot the data and add the fit from the degree-4 polynomial. We set the margins and outer margins in our plot the later plot a title that will be the overall title for two plots that we plot next to each other. The function `matlines()` lets us draw the lines fo the uncertainty bounds in one go.

```{r, eval=FALSE}
# set margins to plot title in margins
par(mfrow = c(1, 2), mar = c(4.5, 4.5, 1, 1), oma = c(0, 0, 4, 0))
plot(wage ~ jitter(age,2), xlim = agelims, cex = 0.5, col = "darkgrey", bty = "n",
     xlab = "age")
# overall plot window title
title("Degree -4 Polynomial ", outer = TRUE)
# line for mean estimate
lines(age.grid, preds$fit, lwd = 2, col = "blue")
# ~95% ci's
matlines(age.grid, se.bands, lwd = 2, col = "blue", lty = 3)
```

```{r, echo=FALSE}
# set margins to plot title in margins
par(mfrow = c(1, 1), mar = c(4.5, 4.5, 1, 1), oma = c(0, 0, 4, 0))
plot(wage ~ jitter(age,2), xlim = agelims, cex = 0.5, col = "darkgrey", bty = "n",
     xlab = "age")
# overall plot window title
title("Degree -4 Polynomial ", outer = TRUE)
# line for mean estimate
lines(age.grid, preds$fit, lwd = 2, col = "blue")
# ~95% ci's
matlines(age.grid, se.bands, lwd = 2, col = "blue", lty = 3)
```

We can compare the orthogonolized polynomials that we saved in the object called `fit` with the polynomials that plain polynomials sved in `fit2`. The difference will substantially $0$. We predict the outcome from the fit with the raw polynomials and take the diffence to the fit with the independent linear combinations of the powers of age.

```{r}
preds2 <- predict(fit2, newdata = list(age = age.grid), se = TRUE)
# average difference
mean(preds$fit - preds2$fit)
# maximum difference
max(abs(preds$fit - preds2$fit))
```

When we have only predictor variable and and its powers we use the `coef()` function to see whether the powers of the variable improve in-sample model fit.

```{r}
fit.5 <- lm(wage ~ poly(age, 5), data = Wage)
coef(summary(fit.5))
```

Preferably, we use the `anova()` and look at the F-test to decide whether in-sample fit improves by including powers of a variable.

```{r}
fit.1 <- lm(wage ~ age, data = Wage)
fit.2 <- lm(wage ~ poly(age, 2), data = Wage)
fit.3 <- lm(wage ~ poly(age, 3), data = Wage)
fit.4 <- lm(wage ~ poly(age, 4), data = Wage)
anova(fit.1, fit.2, fit.3, fit.4, fit.5)
```

We prefer the F-test beacuse it works when we include additional predictors into our model.

```{r}
# comparing models when other predictors are added
fit.1 <- lm(wage ~ education + age, data = Wage)
fit.2 <- lm(wage ~ education + poly(age, 2), data = Wage)
fit.3 <- lm(wage ~ education + poly(age, 3), data = Wage)
anova(fit.1, fit.2, fit.3)
```

With `glm()` we can also fit a polynomial logistic regression. Here, we create a binary variable that is 1 if wage > 250 and 0 otherwise.

```{r}
fit <- glm(I(wage > 250) ~ poly(age, 4), data = Wage, family = binomial)
```

Similar to `lm()` we use the `predict()` function again and also obtain standard errors by setting `se=TRUE`.

Note: If we do **not** set `type="response"` in the `predict()` function, we get the latent $y$ as $X\beta$. We have to send those values through the link function to get predicted probabilities. We do this, so that we can estimate the standard errors on the latent $y$. We then send these through the link function as well. This ensures that our confidence intervals will never be outside the logical $[0, 1]$ interval for probabilities.

```{r}
# predict latent y
preds <- predict(fit, newdata = list(age = age.grid), se = TRUE)

# send latent y through the link function
pfit <- 1 / (1 + exp(-preds$fit))

# error bands calculate on the latent y
se.bands.logit <- cbind(preds$fit + 2 * preds$se.fit, preds$fit - 2 * preds$se.fit)
se.bands <- 1 / (1 + exp(-se.bands.logit))

```

We add the results next to the plot where wage is continuous. With the `points()` function we add the actual data to the plot. The argument `pch="|"` draws a bar as the symbol for each point. Also notice the y-coordinate of each point. In the `plot()` function we set the range of the y-axis with `ylim = c(0, 0.2)` to range from $0$ to $0.2$. If the true outcome is $1$ we want to draw the | at $y=0.2$ and otherwhise at $y=0$. We achieve this with `I((wage > 250)/5)`. Play around to see why.

```{r, eval = FALSE}
plot(I(wage > 250) ~ age, xlim = agelims, type = "n", ylim = c(0, 0.2))
# add data to the plot
points(jitter(age), I((wage > 250)/5) , cex = 1, pch = "|", col = " darkgrey ")
# mean estimate
lines(age.grid, pfit, lwd = 2, col = "blue")
# 95 ci
matlines(age.grid, se.bands, lwd = 2, col = "blue", lty = 3)
```

```{r, echo=FALSE}
par(mfrow = c(1, 2), mar = c(4.5, 4.5, 1, 0), oma = c(0, 0, 4, 0))

# old data
fit <- lm(wage ~ poly(age, 4), data = Wage)
agelims <- range(age)
age.grid <- seq(from = agelims[1], to = agelims[2])
preds <- predict(fit, newdata = list(age = age.grid), se = TRUE)
se.bands <- cbind(preds$fit + 2 * preds$se.fit, preds$fit - 2 * preds$se.fit)

# old plot
plot(jitter(age,2), wage, xlim = agelims, cex = 0.5, col = "darkgrey", bty = "n",
     xlab = "age")
title("Degree -4 Polynomial ", outer = TRUE)
lines(age.grid, preds$fit, lwd = 2, col = "blue")
matlines(age.grid, se.bands, lwd = 2, col = "blue", lty = 3)

# newdata
fit <- glm(I(wage > 250) ~ poly(age, 4), data = Wage, family = binomial)
preds <- predict(fit, newdata = list(age = age.grid), se = TRUE)
pfit <- exp(preds$fit) / (1 + exp(preds$fit))
se.bands.logit <- cbind(preds$fit + 2 * preds$se.fit, preds$fit - 2 * preds$se.fit)
se.bands <- exp(se.bands.logit)/(1 + exp(se.bands.logit))
preds <- predict(fit, newdata = list(age = age.grid), type = "response", se = TRUE)

# second plot
plot(age, I(wage > 250), xlim = agelims, type = "n", ylim = c(0, 0.2), bty = "n")
points(jitter(age), I((wage > 250)/5), cex = 1, pch = "|", col = " darkgrey ")
lines(age.grid, pfit, lwd = 2, col = "blue")
matlines(age.grid, se.bands, lwd = 2, col = "blue", lty = 3)
```

Notice, that the confidence interval becomes very large in the range of the data where we have few data and no $1$'s.


### Step Functions

Instead of using polynomials to create a non-linear prediction, we could also use step functions. With step functions we fit different lines for different data ranges.

We use the `cut()` function to create equally spaced cutpoints in our data. We use the now categorical variable age as predictor in our linear model.

```{r}
# four equally spaced intervals of age
table(cut(age, 4))
# fit the linear regression with the factor variable age that has four categories
fit <- lm(wage ~ cut(age, 4), data = Wage)
# the first category is the baseline.
coef(summary(fit))
```


### Splines

We use the `splines` package to fit splines.

```{r}
library(splines)
```

We first use `bs()` to generate a basis matrix for a polynomial spline and fit a model with knots at age 25, 40 and 60. `bs` will by default fit a cubic spline with the specified number of knots. To deviate from a cubic spline, chanbe the argument `degree` to some other value.

```{r}
fit <- lm(wage ~ bs(age, knots = c(25, 40, 60)), data = Wage)
pred <- predict(fit, newdata = list(age = age.grid), se = TRUE)
par( mfrow = c(1,1))
plot(jitter(age,2), wage, col = "gray", xlab = "age", bty = "n")
lines(age.grid, pred$fit, lwd = 2)
lines(age.grid, pred$fit + 2 * pred$se, lty = "dashed", lwd = 2)
lines(age.grid, pred$fit - 2 * pred$se, lty = "dashed", lwd = 2)
```

The confidence intervals are wide at the end of data range whree the curve is forced off into some direction by only a few observations.

Instead of specifying knots ourselves we can set the argument `df` (degrees of freedom) to some value to set knots at uniform intervals. A polynomial of the power of three uses up four degrees of freedom. One for each power. Where we to fit polynomials in regions of the data we woud use 4 degrees of freedom in each region. Fitting a cubic spline involves three predictors and an intercept = 4. In addition we use a degree of freedom at each knot.

```{r}
dim( bs(age, knots = c(25, 40, 60)) )
dim(bs(age, df = 6))
attr(bs(age, df = 6), "knots")
```


To address the problem of the wide error bands at the ends of the data range, we can fit natural splines. They impose linearity at the end of the range, are better behaved (produce narror error bands) and free up two degrees of freedom. One at each end. We `ns()` for natural splines.

```{r}
plot(jitter(age,2), wage, col = "gray", bty = "n", xlab = "age")
fit2 <- lm(wage ~ ns(age, df = 4), data = Wage)
pred2 <- predict(fit2, newdata = list(age = age.grid), se = TRUE)
lines(age.grid, pred2$fit, col = "red", lwd = 2)
lines(age.grid, pred2$fit + 2 * pred$se, lty = "dashed", lwd = 2, col = "red")
lines(age.grid, pred2$fit - 2 * pred$se, lty = "dashed", lwd = 2, col = "red")
```

### Local Regression

One can also account for non-linearity by using local regression. We estimate the regression curve only in the near neighborhood of the the current point. Smoothness is achieved by moving along the data range gradually whereby the intervals overlap. Like k-nearest neighbors local regression can does not work well in high dimensions (many predictors) because the distances between points become very large.

Here, wer use the `loess()` smoother for local regression. The argument `span` specifies the range of the data that is being used. The narrower the span the more flexible and wiggly the line will become.

```{r}
plot(jitter(age,2), wage, xlim = agelims, cex = 0.5, col = "darkgrey", bty = "n", xlab = "age")
title(" Local Regression ")
fit <- loess(wage ~ age, span = 0.2, data = Wage)
fit2 <- loess(wage ~ age, span = 0.5, data = Wage)
lines(age.grid, predict(fit, data.frame(age = age.grid)), col = "red", lwd = 2)
lines(age.grid, predict(fit2, data.frame(age = age.grid)), col = "blue", lwd = 2)
legend("topright", legend = c("Span=0.2", "Span=0.5"), col = c("red", "blue"), 
       lty = 1, lwd = 2, cex = 0.8, bty = "n")
```


### Genearlized Additive Models (GAMs)

With GAMs we can flexibly combine the mehtods we introduced above for multible predictors.

For splines, we can fit a GAM with `lm()` when an appropriate basis functions can be used.

```{r}
gam1 <- lm(wage ~ bs(year, 4) + bs(age, 5) + education, data = Wage)
```

However, the `gam` package offers a general solution to fitting GAMs and is especially useful when splines cannot be easily expressed in terms of basis functions. Here, we fit natural splines on age and year. Something that could not easily be expressed as a basis function.

```{r}
library(gam)
# gam with natural splines for year and age
gam.m3 <- gam(wage ~ ns(year, 4) + ns(age, 5) + education, data = Wage)
```

We can use the `plot()` function as well as `plot.gam()` to plot our results. The difference between cubic splines and natural splines fit to age and education is very small.


```{r}
par(mfrow = c(1, 3), mar = c(4.5, 4.5, 1, 1), oma = c(0, 0, 4, 0))
# plot function behaves similar to plot.gam
plot(gam.m3, se = TRUE, col = "blue")
title("Natural splines for age and education", outer = TRUE)

par(mfrow = c(1, 3), mar = c(4.5, 4.5, 1, 1), oma = c(0, 0, 4, 0))
plot.gam(gam1, se = TRUE, col = "red")
title("Cubic splines for age and education", outer = TRUE)
```

We can use `anova()` to find the best performing model.

```{r}
gam.m1 <- gam(wage ~ ns(age, 5) + education, data = Wage)
gam.m2 <- gam(wage ~ year + ns(age, 5) + education, data = Wage)
anova(gam.m1, gam.m2, gam.m3, test = "F")
```

We can see that looking at the `summary()` function would not have been enough to determine whether a natural spline on year is appropriate. The coeffiecient is highly significant. However, the F-test reveals that the model is not significantly better than the model without a natural spline for year.

```{r}
summary(gam.m3)
```

We can make predictions using `predict()` just as we did before.

```{r}
preds <- predict(gam.m2, newdata = Wage)
```

The `gam()` function also allows fitting logistic regression GAM with the `family = binomial` argument.

```{r}
gam.lr <- gam(I(wage > 250) ~ year + ns(age, df = 5) + education, family = binomial, data = Wage)
par(mfrow = c(1, 3))
plot(gam.lr, se = TRUE, col = "green")
```

Looking at the education variable, we see that the first category has extremely wide error bands. We check a cross-table of education against the indicator of whether age > 250. We see that age > 250 is never true for lowest category. 

```{r}
table(education, I(wage > 250))
```

Based on this we exclude the lowest education category from our model.

```{r}
gam.lr.s <- gam(I(wage > 250) ~ year + ns(age, df = 5) + education, family = binomial, data = Wage, subset = (education != "1. < HS Grad"))
plot(gam.lr.s, se = TRUE, col = "green")
```


### Exercises

#### Q1

In this exercise, you will further analyze the `Wage` dataset coming with the `ISLR` package.

1. Perform polynomial regression to predict `wage` using `age`. Use cross-validation to select the optimal degree dd for the polynomial. What degree was chosen, and how does this compare to the results of hypothesis testing using `ANOVA`? Make a plot of the resulting polynomial fit to the data.
2. Fit a step function to predict `wage` using `age`, and perform cross-validation to choose the optimal number of cuts. Make a plot of the fit obtained.

#### Q2

The `Wage` data set contains a number of other features that we haven't yet covered, such as marital status (`maritl`), job class (`jobclass`), and others. Explore the relationships between some of these other predictors and `wage`, and use non-linear fitting techniques in order to fit flexible models to the data. Create plots of the results obtained, and write a summary of your findings.

#### Q3

This question uses the variables `dis` (the weighted mean of distances to five Boston employment centers) and `nox` (nitrogen oxides concentration in parts per 10 million) from the `Boston` data available as part of the `MASS` package. We will treat `dis` as the predictor and `nox` as the response.

1. Use the `poly()` function to fit a cubic polynomial regression to predict `nox` using `dis`. Report the regression output, and plot the resulting data and polynomial fits.
2. Plot the polynomial fits for a range of different polynomial degrees (say, from 1 to 10), and report the associated residual sum of squares.
3. Perform cross-validation or another approach to select the optimal degree for the polynomial, and explain your results.
4. Use the `bs()` function to fit a regression spline to predict `nox` using `dis`. Report the output for the fit using four degrees of freedom. How did you choose the knots? Plot the resulting fit.
5. Now fit a regression spline for a range of degrees of freedom, and plot the resulting fits and report the resulting RSS. Describe the results obtained.

#### Q4

This question relates to the `College` dataset from the `ISLR` package.

1. Split the data into a training set and a test set. Using out-of-state tuition as the response and the other variables as the predictors, perform forward stepwise selection on the training set in order to identify a satisfactory model that uses just a subset of the predictors.
2. Fit a GAM on the training data, using out-of-state tuition as the response and the features selected in the previous step as the predictors. Plot the results, and explain your findings.
3. Evaluate the model obtained on the test set, and explain the results obtained.
4. For which variables, if any, is there evidence of a non-linear relationship with the response?
