---
title: "Lab 1  -- Solutions"
author: "Philipp Broniecki and Lucas Leemann -- Machine Learning 1K"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##### R Introduction   (based on Leemann and Mykhaylov, PUBLG100)

2. Clear the workspace and set the working directory to your `StatisticalLearning` folder.

```{r, eval=FALSE}
# clear workspace
rm( list = ls() )
```

We set the working directory with `setwd()`. You can find a detailed description [here](https://uclspp.github.io/PUBLG100/faq/setwd.html).

```{r, eval=FALSE}
setwd("C:/some folder/some folder/StatisticalLearning")
```

3. Load the High School and Beyond dataset. Remember to load any necessary packages.

```{r}
setwd("C:/Users/phili/Documents/ML2017.io")
library(readxl)
df <- read_excel("./data/hsb2.xlsx")
```

4. Calculate the final score for each student by averaging the `read`, `write`, `math`, `science`, and `socst` scores and save it in a column called  final_score.

We use the `apply()` function. Check its helpfiles by typing `?apply`. We pass a subset of our dataset only including the 5 variables to `apply()`. The second argument `1` tells `apply()` to work rowwise. The third argument tells it to take the mean.

```{r}
df$final_score <- apply(df[, c("read", "write", "math", "science", "socst")], 1, mean)
```

5. Calculate the mean, median and mode for the final_score.

```{r}
mean(df$final_score)
median(df$final_score)
```

There is no function for the mode. So we look at a frequency table of the `final_score` using the `table()` function. We put this inside the `which.max()` function which returns the index of the most frequent score. Finally, we index the frequency table with the id we obtained. 

```{r}
id.of.mode <- which.max(table(df$final_score))
table(df$final_score)[id.of.mode]
```

6. Create a factor variable called school_type from schtyp using the following codes:
    - 1 = Public schools
    - 2 = Private schools
    
```{r}
df$schtyp <- factor(df$schtyp, labels = c("Public schools", "Private schools"))
```

7. How many students are from private schools and how many are from public schools?

We again use the `table()` function to produce a frequency table.

```{r}
table(df$schtyp)
```

8. Calculate the variance and standard deviation for `final_score` from each school type.

Do do this we index the data with the square brackets. Inside we put a condition. Only the data where the condition is true will be evaluated. The double equal signs mean *is the same*

```{r}
# variance
var(df$final_score[df$schtyp == "Private schools"])
var(df$final_score[df$schtyp == "Public schools"])

# standard deviation
sd(df$final_score[df$schtyp == "Private schools"])
sd(df$final_score[df$schtyp == "Public schools"])
```

9. Find out the ID of the students with the highest and lowest `final_score` from each school type.

We use the `which.min()` and `which.max()` functions again to get the index numbers of the correct students.

```{r}
# best student's id from private school
df$id[ which.max(df$final_score[ df$schtyp == "Private schools" ]) ]

# best student's id from public school
df$id[ which.max(df$final_score[ df$schtyp == "Private schools" ]) ]

# worst student's id from private school
df$id[ which.min(df$final_score[df$schtyp == "Private schools"])]

# worst student's id from private school
df$id[ which.min(df$final_score[df$schtyp == "Public schools"])]
```

10. Find out the 20th, 40th, 60th and 80th percentiles of final_score.

The `quantile()` function allows us to do this easily.

```{r}
quantile(df$final_score, c(.2, .4, .6, .8))
```

12. Create box plot for `final_score` grouped by the school_type factor variable to show the difference between `final_score` at public schools vs. private schools.

The `plot()` function knows that a boxplot is appropriate becuase `schtyp` is a factor variable.

```{r}
plot(final_score ~ schtyp, data = df, bty = "l", col = c("#AA4455", "#114477"), 
     xlab = "type of school", ylab = "final score", frame.plot = FALSE)
```

```{r, echo=FALSE}
rm(list = ls())
detach(package:readxl, unload = TRUE)
```