---
title: "Lab 2  -- Solutions"
author: "Philipp Broniecki and Lucas Leemann -- Machine Learning 1K"
---

### Solution Lab 2

1. Load the immigration dataset dataset

```{r}
# clear workspace
rm(list=ls())

# read in various datasets and combine them 
immi <- read.csv("http://uclspp.github.io/PUBLG100/data/communities_immig.csv")

```


2. Rename the `PctImmigRec5` variable to `RecentImmigration5`.

Here, we use `select()` from `dplyr`. Sometimes different packages provide packages with similar names. Both `dplyr` and `Zelig` have functions called `select()` for example. This can cause conflicts. If we have a package installed, we can call a function from it directly without loading the function of that package into the current namespace. The following code does this. We also have to merge some data sets.


```{r}
# load original communities data set
com <- read.csv("C:/Users/phili/Dropbox/Essex 2017 ML/Day 2/Lab/communities.csv")

# load data set on employment
empl <- read.csv("http://uclspp.github.io/PUBLG100/data/communities_employment.csv")

# merge data sets
# com and empl
com <- merge(com, empl, by = c("state", "communityname"))
# merge immi to the data sets
com <- merge(com, immi, by = c("state", "communityname"))
# remove the now unecessary data sets immi and empl
rm(immi, empl)

com <- dplyr::select(com,
                     community = communityname,
                     unemploymentrate = PctUnemployed.y,
                     nohighschool = PctNotHSGrad,
                     white = racePctWhite,
                     recentimmigration5 = PctImmigRec5
                     )
```

3. Estimate a model explaining the relationship between unemployment rate and recent immigration over the past 5 years using the variable `RecentImmigration5`.

```{r}
m_immi <- lm(unemploymentrate ~ recentimmigration5, data = com)
summary(m_immi)
```

4. Plot the regression line of the model.

We draw a scatterplot using the `plot()` function and `abline()` to draw the regression line.

```{r}
# the scatterplot
plot(com$unemploymentrate ~ com$recentimmigration5,
     xlab = "Recent Immigration",
     ylab = "Unemployment Rate",
     bty = "n", pch = 16)

# add the regression line
abline(m_immi, col = "red", lwd = 2)
```

5. How does this model compare to the models we estimated in the seminar with NoHighSchool and Minority as independent variables? Present your findings by comparing the output of all three models in side-by-side tables using the texreg package.

```{r}
m_edu <- lm(unemploymentrate ~ nohighschool, data = com)

# minority percentage
com$minority <- 1 - com$white
m_minority <- lm(unemploymentrate ~ minority, data = com)

texreg::screenreg(list(m_edu, m_minority, m_immi))
```

6. Save your model comparison table to a Microsoft Word document (or another format if you don’t use Word).

```{r}
texreg::htmlreg(list(m_edu, m_minority, m_immi), file = "Lab2_model_comparison.doc")
```

7. Generate predicted values from the fitted model with RecentImmigration5 and plot the confidence interval using `Zelig`.

```{r}
library(Zelig)
z.out <- zelig(unemploymentrate ~ recentimmigration5, data = com, model = "ls")
x.out <- setx(z.out, recentimmigration5 = seq(0, 1, 0.1))
s.out <- sim(z.out, x = x.out, n=10000)
```

8. Save all the plots as graphics files.
```{r}
png("plot.png")
ci.plot(s.out, xlab = "Recent Immigration (last 5 years)", ci = 95)
dev.off()
```