---
title: "Solution Day 3 Lab - Classification"
author: "Philipp Broniecki and Lucas Leemann -- Machine Learning 1K"
---

You will need to load the core library for the course textbook:

```{r}
# clear workspace
rm(list=ls())

library(ISLR)
```

#### Q1

In this problem, you will develop a model to predict whether a given car gets high or low gas mileage based on the `Auto` dataset from the `ISLR` package.

1. Create a binary variable, `mpg01`, that contains a 1 if `mpg` contains a value above its median, and a 0 if `mpg` contains a value below its median. You can compute the median using the `median()` function. Note you may find it helpful to use the `data.frame()` function to create a single data set containing both `mpg01` and the other `Auto` variables. Type `?Auto` to get the codebook of the dataset.

```{r}
library(ISLR)
summary(Auto)

attach(Auto)
mpg01 <- ifelse( mpg > median(mpg), yes = 1, no = 0)
Auto <- data.frame(Auto, mpg01)
```

2. Explore the data graphically in order to investigate the association between `mpg01` and the other features. Which of the other features seem most likely to be useful in predicting `mpg01`? Scatterplots and boxplots may be useful tools to answer this question. Describe your findings.

```{r}
cor(Auto[,-9])

pairs(Auto) # doesn't work well since mpg01 is 0 or 1
```

**cylinders, weight, displacement, horsepower (and mpg itself)**

3. Split the data into a training set and a test set.

First, we normalize our varibles
```{r}
Auto <- data.frame(mpg01, apply(cbind(cylinders, weight, displacement, horsepower, acceleration), 
                                2, scale), year)
```

We use the `%%` operator to split the data set into observations with even and uneven years. Check [here]("http://www.statmethods.net/management/operators.html") for a description of different R operators.

```{r}
train <-  (year %% 2 == 0) # if the year is even (%%)
test <-  !train
Auto.train <-  Auto[train,]
Auto.test <-  Auto[test,]
mpg01.test <-  mpg01[test]
```

4. Perform LDA on the training data in order to predict `mpg01` using the variables that seemed most associated with `mpg01` in (b). What is the test error of the model obtained?

```{r}
# LDA
library(MASS)
lda.fit <-  lda(mpg01 ~ cylinders + weight + displacement + horsepower,
              data = Auto, subset = train)
lda.pred <-  predict(lda.fit, Auto.test)
mean(lda.pred$class != mpg01.test)
```

**12.6% test error rate.**

5. Perform logistic regression on the training data in order to predict `mpg01` using the variables that seemed most associated with `mpg01` in (b). What is the test error of the model obtained?

```{r}
# Logistic regression
glm.fit <-  glm(mpg01 ~ cylinders + weight + displacement + horsepower,
              data = Auto,
              family = binomial,
              subset = train)
glm.probs <-  predict(glm.fit, Auto.test, type = "response")
glm.pred <-  rep(0, length(glm.probs))
glm.pred[glm.probs > 0.5] <- 1
mean(glm.pred != mpg01.test)
```

**12.1% test error rate.**

6. Perform KNN on the training data, with several values of K, in order to predict `mpg01`. Use only the variables that seemed most associated with `mpg01` in (b). What test errors do you obtain? Which value of K seems to perform the best on this data set?

```{r}
library(class)
train.X <-  cbind(cylinders, weight, displacement, horsepower)[train,]
test.X <-  cbind(cylinders, weight, displacement, horsepower)[test,]
train.mpg01 <-  mpg01[train]
set.seed(1)

# KNN (k=1)
knn.pred <-  knn(train.X, test.X, train.mpg01, k = 1)
mean(knn.pred != mpg01.test)

# KNN (k=10)
knn.pred <-  knn(train.X, test.X, train.mpg01, k = 10)
mean(knn.pred != mpg01.test)

# KNN (k=100)
knn.pred <-  knn(train.X, test.X, train.mpg01, k = 100)
mean(knn.pred != mpg01.test)
```

**k=1, 15.4% test error rate. k=10, 16.5% test error rate. k=100, 14.3% test error rate. K of 100 seems to perform the best. 100 nearest neighbors.**

7. We can find the optimal value for K by trying all K from 1 to 200. We will run a loop and save the test error for every value of K.

```{r}
K.collector <- rep(NA, 200)
for (k.try in 1:200){
  knn.pred <-  knn(train.X, test.X, train.mpg01, k = k.try)
  K.collector[k.try] <- mean(knn.pred != mpg01.test)
}
x.k <- c(1:200)

plot(x.k,K.collector, type="o", pch=19, cex=0.5, main="K ")
```

It is hard to see the optimal value of K for us. We can inspect the error ourselves:

```{r}
which.min(K.collector)
```

**Here, it seems to be the case that K=4 is the best. We will see tomorrow how we can ensure that this does not just depend on the specific trianing and test data we created here.**


#### Q2

This problem involves writing functions.

1. Write a function, `Power()`, that prints out the result of raising 2 to the 3rd power. In other words, your function should compute $2^3$ and print out the results.

Hint: Recall that $x^a$ raises x to the power a. Use the `print()` function to output the result.

```{r}
Power <-  function() {
  2^3
}
print(Power())
```

2. Create a new function, `Power2()`, that allows you to pass any two numbers, `x` and `a`, and prints out the value of `x^a`. You can do this by beginning your function with the line:

```{r, eval=FALSE}
Power2 <- function(x,a) {
```

You should be able to call your function by entering, for instance,

```{r, eval=FALSE}
Power2(3,8)
```

on the command line. This should output the value of $3^8$, namely, 6,561.

```{r}
Power2 <-  function(x, a) {
  x^a
}
Power2(3,8)
```

3. Using the `Power2()` function that you just wrote, compute $10^3$, $8^{17}$, and $131^3$.

```{r}
Power2(10, 3)

Power2(8, 17)

Power2(131, 3)
```

4. Now create a new function, `Power3()`, that actually returns the result `x^a` as an R object, rather than simply printing it to the screen. That is, if you store the value `x^a` in an object called result within your function, then you can simply `return()` this result, using the following line:

```{r, eval=FALSE}
return(result)
```

The line above should be the last line in your function, before the `}` symbol.

```{r}
Power3 <-  function(x, a) {
  result <-  x^a
  return(result)
}
```

5. Now using the `Power3()` function, create a plot of $f(x)=x^2$. The $x$-axis should display a range of integers from 1 to 10, and the $y$-axis should display $x^2$. Label the axes appropriately, and use an appropriate title for the figure. Consider displaying either the $x$-axis, the $y$-axis, or both on the log-scale. You can do this by using `log="x"`, `log="y"`, or `log="xy"` as arguments to the `plot()` function.

```{r}
x <-  1:10
plot(x, Power3(x, 2),  
     log="xy", ylab="Log of y = x^2", xlab="Log of x", 
     main="Log of x^2 versus Log of x", bty = "n")
```

6. Create a function, `PlotPower()`, that allows you to create a plot of `x` against `x^a` for a fixed `a` and for a range of values of `x`. For instance, if you call

```{r, eval=FALSE}
PlotPower(1:10, 3)
```

then a plot should be created with an $x$-axis taking on values $1,2,...,10$, and a $y$-axis taking on values $1^3,2^3,...,10^3$.

```{r}
PlotPower <-  function(x, a) {
  plot(x, Power3(x, a), bty = "n")
}
PlotPower(1:10, 3)
```
